Variability as a better characterization of Shannon entropy

### 2019-12-28 Articles/2019-ShannonEntropy/Submissions/AJP/AJP-ShannonEntropyIsVariability.tex
Updated introduction and citations

The Shannon entropy $-\sum p_i \log p_i$ is well established in such disparate fields as computer science, physics and ecology. It is most often introduced in physics through statistical mechanics, and associated with vague notions, such as uncertainty, knowledge, lack of knowledge and disorder. These prevent a proper understanding of the concept and often lead to confusion.

This paper aims to give a crisp characterization of the Shannon entropy that is intuitive and precise. The general idea is that it measures the variability of the elements within a distribution, which is a general concept applicable to many branches of science, and it is independent of the notion of entropy of thermodynamics and statistical mechanics. Therefore we will use the term \emph{``Shannon variability,''} leaving ``entropy'' to the physical concept. We will show that the expression $- \sum_i p_i \log p_i$ is the only linear indicator of variability. It measures the number of questions one must ask to identify an element of the distribution, linking its use to information theory. We will study how the expression works over continuous variables, and show how phase space is special as it leaves the Shannon variability invariant under change of coordinates. We will turn to statistical mechanics and apply Shannon variability in different ways to recover connections to the Boltzmann, Gibbs and von Neumann entropies.

While some details presented here may be well known in a particular community, we find that they may be new to another. Therefore we assume little knowledge of the subject from the reader and include all details that we feel are important to properly understand the subject without confusion, including two standard derivations for the formula and examples that are common in computer science but rarely used in physics. We will briefly touch on a few of the common misconceptions.

>>>>

The Shannon entropy $-\sum p_i \log p_i$ is well established in such disparate fields as computer science, physics\cite{Jaynes1957-1, Grandy} and ecology, yet it is still surrounded by a great deal of confusion.\cite{Swendsen, KishFerry} For example, in statistical mechanics it is often introduced and associated with vague notions (such as uncertainty, knowledge, lack of knowledge and disorder) which prevent a proper understanding of the concept.\cite{Styer, Ben-Naim} In this work we concentrate on giving a better characterization of the Shannon entropy and its relation to statistical mechanics by addressing what we believe are the two main problems; we do not discuss the relationship to the thermodynamic entropy, which is separate concept.

The first problem is that the formula is attached to different concepts that are specific to each field. To disentangle this issue we need to provide a characterization of the Shannon entropy that satisfies the following requirements:
\begin{itemize}
	\item it is defined in a general setting that is crisp and intuitive
	\item it is precise enough so that the formula itself can be derived
	\item it recovers the specific meaning when applied to each field.
\end{itemize}
In this paper we show that the idea of variability of the elements within a distribution, how diverse the objects are from each other within the collection, can serve as such a general concept. We will therefore use the term \emph{``Shannon variability,''} leaving ``entropy'' to the physical concept. We will show that the expression $- \sum_i p_i \log p_i$ is the only linear indicator of variability and it measures the number of questions one must ask to identify an element of the distribution, linking its use to information theory. The variability of messages within a communication channel will correspond to its information entropy as used in communication theory; the variability of microstates within a macrostate will correspond to the Gibbs entropy used in statistical mechanics; the variability of the state of a single particle within a microstate will correspond to the Boltzmann entropy.

The second problem is that the Shannon formula presents some peculiarity when applied over continuous variables that is often the source of confusion. Namely, it is generally linked to a choice of unit. Yet, the physical state spaces, in both classical and quantum mechanics, provide additional structure that make the Shannon variability coordinate-independent. This is seldom pointed out, and it is critical to form a proper connection to the Boltzmann, Gibbs and von Neumann entropies.

While some details presented here may be well known in a particular community, we find that they may be new to another. Therefore we assume little knowledge of the subject from the reader and include all details that we feel are important to properly understand the subject without confusion, including two standard derivations for the formula and examples that are common in computer science but rarely used in physics. We will briefly touch on a few of the common misconceptions and show how they are resolved.
